---
title: "Week 6 live code"
output: html_document
date: "2024-09-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Example code provided during live lecture

**Step 1: Generate the Dataset We’ll create two samples:**
Sample 1: Normally distributed data.   
Sample 2: Slightly right-skewed data from a Chi-square distribution with the same mean as the normal distribution.

**Step 2: Perform Statistical Tests**

• Wilcoxon rank-sum test: A non-parametric test that does not assume normality.

• Welch’s two-sample t-test: A t-test that does not assume equal variances between the two samples.

**Step 3: Visualize the Data**

We will use a combination of histograms and density plots to visualize the differences between the two samples

```{r}
# Load necessary libraries

library(ggplot2)

# Set seed for reproducibility
set.seed(100)

# Parameters
n <- 30  # Moderate sample size
mu <- 8   # Mean for the normal distribution
sigma <- 2  # Standard deviation for the normal distribution
df <- 10  # Degrees of freedom for chi-square distribution (moderate skewness)


# Generate sample 1 from a normal distribution
sample1 <- data.frame(x =  rnorm(n, mean = mu, sd = sigma), distribution = "Normal Distribution")

# Generate sample 2 from a chi-square distribution, shifted to have the same mean as the normal
sample2 <- data.frame(x =  rchisq(n, df = df), distribution = "Chi-square Distribution")

combined_df <- rbind(sample1, sample2)

## generate some summary stats for the two samples 
combined_df |> group_by(distribution) |> 
 summarise(mean = mean(x), 
          variance = sd(x)^2, 
          median = median(x), 
            n = n())
```




```{r}
# Perform Wilcoxon rank-sum test (non-parametric)
wilcox_test <- wilcox.test(sample1$x, sample2$x)

# Print test results
print(wilcox_test)
```

```{r}
# Perform Welch's two-sample t-test (assumes unequal variances)
welch_test <- t.test(sample1$x, sample2$x, var.equal = FALSE)

# Print test results
print(welch_test)
```

Note that the echo = FALSE parameter was added to the code chunk to prevent printing of the R code that generated the plot.
