---
title: "Week 4"
format: 
  html:
    code-tools: true
    embed-resources: true
---

```{r, message=FALSE}
library(tidyverse)
```

## One sample t-test

### Beer contents

```{r}
y = c(374.8, 375.0, 375.3, 374.8, 374.4, 374.9)
```

```{r}
df = data.frame(y)
set.seed(124)
p1 = ggplot(df, aes(x = "", y = y)) +
  geom_boxplot(alpha = 0.5, coef = 10) + 
  geom_dotplot(binaxis = 'y', 
               stackdir = 'center') + 
  geom_hline(yintercept = 375, 
             colour = "blue",
             linetype = "dashed") + 
  labs(y = "Beer volume (ml)", x = "") +
  theme_bw(base_size = 24) + 
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
p1
```

```{r}
x = c(374.8, 375.0, 375.3, 374.8, 374.4, 374.9)
mean(x)
sd(x)
t.test(x, mu = 375, alternative = "less")
n = length(x)
t0 = ( mean(x)-375 ) / ( sd(x)/sqrt(n) ) 
t0
pval = pt(t0, n-1)
pval
```

## Two sample t-test

### Equal variance

```{r}
non_smokers = c(25, 25, 27, 44, 30, 67, 
                53, 53, 52, 60, 28)
smokers =  c(27, 29, 37, 36, 46, 82, 
             57, 80, 61, 59, 43)
dat = data.frame(
  platelets = c(non_smokers, smokers),
  status = c(rep("Non smokers", 
                 length(non_smokers)),
             rep("Smokers", 
                 length(smokers)))
)
sum = dat |> 
  group_by(status) |> 
  summarise(Mean = mean(platelets),
            SD = sd(platelets), 
            n = n())
```

```{r}
knitr::kable(sum, 
             format = "html", 
             digits = 1)
gt::gt(sum) |> 
  gt::fmt_number(columns = c("Mean","SD"), decimals = 1)
```

```{r}
ggplot(dat) + aes(x = status, y = platelets) +
  geom_boxplot() + 
  geom_jitter(width = 0.15, size = 3, colour = "blue") + 
  labs(x = "", y = "Blood platelet\naggregation")
```

```{r}
t.test(smokers, non_smokers,
       alternative = "two.sided",
       var.equal = TRUE)
```

```{r}
nS = length(smokers)
nN = length(non_smokers)
sS = sd(smokers)
sN = sd(non_smokers)
sP = sqrt(((nS - 1) * sS^2 + (nN - 1) * sN^2)/(nS + nN - 2))
xbarS = mean(smokers)
xbarN = mean(non_smokers)
deg_free = nS+nN-2
t0 = (xbarS - xbarN) / ( sP * sqrt(1/nS+1/nN) ) 
p_val = 2*(1-pt(abs(t0),deg_free))
c(t0, p_val)
```

### Unequal varaiances

```{r}
sum |> gt::gt() |> gt::fmt_number(columns = 2:3, decimals = 1)
```

```{r}
t.test(smokers, non_smokers, alternative = "two.sided")
```

## Paired t-test

```{r}
before = c(25, 25, 27, 44, 30, 67, 53, 
           53, 52, 60, 28)
after =  c(27, 29, 37, 36, 46, 82, 57, 
           80, 61, 59, 43)
df = data.frame(before, after,
  difference = after - before)
df
```

```{r}
df |> 
  summarise(across(.cols = c(before, after, difference),
                   .fns = list(Mean = mean, 
                               SD = sd, 
                               n = length))) |> 
  pivot_longer(cols = everything(),
               names_sep = "_",
               names_to = c("time", ".value"))
```

```{r}
p = ggplot(df) +
  aes(x = "", y = difference) + 
  geom_boxplot() +
  geom_dotplot(binaxis = "y", stackdir = "center") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = 'Difference in blood platelet levels')+
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p
```

```{r}
t.test(df$after, df$before, paired = TRUE)
t.test(df$difference)
```

## Critical values, rejection regions and confidence intervals

## Beer contents

```{r}
x = c(374.8, 375.0, 375.3, 374.8, 374.4, 374.9)
mean(x)
sd(x)
```

```{r}
df = data.frame(x)
set.seed(124)
ggplot(df, aes(x = "", y = x)) +
  geom_boxplot(alpha = 0.5, coef = 10) + 
  geom_dotplot(binaxis = 'y', 
               stackdir = 'center') + 
  geom_hline(yintercept = 375, 
             colour = "blue",
             linetype = "dashed") + 
  labs(y = "Beer volume (ml)", x = "") +
  theme_bw(base_size = 34) + 
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
```

## Two-sided

```{r}
xbar = mean(x)
xbar
se = sd(x)/sqrt(6)
se
```

```{r}
discrep=abs(xbar-375)
discrep 
```

Comparing the size of the discrepancy to the standard error:

```{r}
discrep/se
```

To reject the null hypothesis (at the 5% level of significance) we would want that ratio to be at least

```{r}
qt(0.975, 5)
```

95% Confidence interval

```{r}
c_95 = qt(0.975,5)
c_95
xbar + c(-1,1) * c_95 * se
```

99% confidence interval

```{r}
c_99 = qt(0.995,5)
c_99
xbar + c(-1,1)*c_99*se
```

Using `t.test()`

```{r}
t.test(x, mu = 375)
t.test(x, mu = 375, conf.level = 0.99)
```

## One sided

```{r}
c_05 = qt(.95, 5)
c_05
c(xbar - 375, se)
c_01 = qt(.99, 5)
c_01
c_05 = qt(.95, 5)
xbar + c_05 * se
c(-Inf, xbar + c_05*se)
c_01 = qt(.99, 5)
c(-Inf, xbar + c_01 * se)

```

```{r}
t.test(x, mu = 375, alternative = "less")
t.test(x, mu = 375, alternative = "less", conf.level = 0.99)
t.test(x, mu = 375, conf.level = 1 - 0.3177)
t.test(x, mu = 375, alternative = "less", conf.level = 1-0.1589)
```

## Smoking and blood platelet aggregation

```{r}
before = c(25, 25, 27, 44, 30, 67, 53, 53, 52, 60, 28)
after =  c(27, 29, 37, 36, 46, 82, 57, 80, 61, 59, 43)
df = data.frame(before, after, difference = after-before)
```

```{r}
p = ggplot(df, aes(x="", y=difference)) +
geom_boxplot() +
geom_dotplot(binaxis = "y", stackdir = "center") +
theme_classic(base_size = 24) +
geom_hline(yintercept = 0, linetype='dashed') +
labs(y = 'Difference in blood platelet levels')+
theme(axis.title.x=element_blank(),
axis.text.x=element_blank(),
axis.ticks.x=element_blank())
p

```

```{r}
n = length(df$difference)
dbar = mean(df$difference)
s_d = sd(df$difference)
t0 = dbar/(s_d/sqrt(n))
c(n, dbar, s_d, t0) |>
  round(2)
alpha = 0.05
qt(1-alpha/2, n - 1)
```

```{r}
(s_d = sd(df$difference))
n=nrow(df); mu0=0
(crit_val=qt(0.975,n-1))
rrlower=mu0-crit_val*s_d/sqrt(n)
rrupper=mu0+crit_val*s_d/sqrt(n)
c(rrlower,rrupper) |> round(2)
```
