---
title: "Rats"
output: html_document
date: "2024-08-26"
---

## Does the biochemical substance have an inhibitive effect on muscular growth?

### Paired $t$-tests approach

```{r}
library(tidyverse)
library(dplyr)
# rat holds the dataframe for the bio vs the normal leg. 
# d is the difference, weight [g]
rat = data.frame(
  bio = c(1.7, 2.0, 1.7, 1.5, 1.6, 
          2.4, 2.3, 2.4, 2.4, 2.6),
  pla = c(2.1, 1.8, 2.2, 2.2, 1.5, 
          2.9, 2.9, 2.4, 2.6, 2.5)
) |> mutate(d = pla - bio)
rat |> summarise(mean = mean(d), sd = sd(d))
```



```{r}
rat |> ggplot() + 
  aes(x = "", y = d) + 
  geom_boxplot() + 
  geom_dotplot(
    binaxis = "y", 
    stackdir = "center") + 
  labs(x = "", 
       y = "Difference\n(placebo - biochemical)")
```

\textbf{Workflow}

Hypothesis: $H_0: \mu_d = 0$ \textbf{vs} $H_1: \mu_d > 0$
    
Assumptions: $D_i$ are independent and identically distributed (iid) $\mathcal{N}(\mu, \sigma^2)$.
    
Test statistic:
$$ T = \frac{\bar{D} - \mu_d}{S_d/\sqrt{n}}. \text{ Under } H_0, T \sim t_{n-1}$$
    
Observed test statistic:

Since they ask for an "inhibitive effect", we use "greater"
```{r}
t.test(x = rat$d, alternative = "greater")
```

$$ t_0 = \frac{0.25}{0.33/\sqrt{10}} = 2.39$$
    
p-value:
$$P(t_9 \geq 2.39) = 0.02$$
    
Conclusion: The p-value of 0.02 is less than 0.05, therefore we reject the null hypothesis at the $5\%$ level of significance and conclude that the biochemical substance does inhibit muscle growth.


## Critical value approach

Given $\alpha = 0.05$, then the critical value $c$ is:
```{r}
qt(p = 0.95, df = 9) # c
```

This means that if $t_0 > c \Rightarrow$ reject $H_0$ given that the rejection region is in the upper tail.

Why is the rejection region only in the upper tail of the t distribution?  
Because:"inhibitive effect" => we use "greater"


## Rejection region approach

For $H_1: \mu_d > \mu_0$ we can define a rejection region using our test statistic:
$$t_0 = \frac{\bar{D} - \mu_0}{s_d / \sqrt{n}} \geq c$$

Noting that,
$$ \alpha = P\left( \frac{\bar{D} - \mu_0}{S_d / \sqrt{n}} \geq c \right) = P\left( \bar{D} \geq \mu_0 + c \frac{s_d}{\sqrt{n}} \right)$$

we define a rejection region on the measurement scale as:
$$\left\{ \bar{D}: \bar{D} \geq \mu_0 + c \frac{s_d}{\sqrt{n}} \right\}$$

Calculations:
```{r}
n = nrow(rat)
qt(0.95, df = 9) * sd(rat$d)/sqrt(n)
```

$$\mu_0 + cs_d / \sqrt{n} = 1.833 \times 0.33 / \sqrt{10} = 0.19$$

Rejection region on the measurement scale:
$$\left\{ \bar{d}: \bar{d} \geq 0.19 \right\}$$

Hence, we would reject the null hypothesis when we observe a sample mean difference greater than 0.19 grams.

In the present case, the sample mean difference was 0.25 grams, so we reject the null hypothesis.


# Sign Test

```{r}
rat |> mutate(
  pos_d = d > 0
) |> arrange(d)
```

```{r}
# number of positive differences (TRUE = 1, FALSE = 0)
sum(rat$d > 0) 

# number of non-zero differences
sum(rat$d != 0)
```

## Workflow

Let $p_+$ be the probability of a positive difference between the treated leg and the placebo leg (difference = placebo - treatment).

\begin{itemize}
\item \textbf{Hypothesis:} $H_0: p_+ = \frac{1}{2}$ \textbf{vs} $H_1: p_+ > \frac{1}{2}$
    
\item \textbf{Assumptions:} Differences, $D_i$, are independent.
    
\item \textbf{Test statistic:} Let $T$ be the number of positive differences out of the 9 non-zero differences. Under $H_0$, $T \sim B(9, \frac{1}{2})$. I.e. under $H_0$, $T$ follows a binomial distribution with $n = 9$ and $p = 0.5$.
    
\item \textbf{Observed test statistic:} We observed $t_0 = 6$ positive differences in the sample.
    
\item \textbf{p-value:} Probability of a test statistic as or more extreme than we observed,
$$P(T \geq 6) = 1 - P(T \leq 5) = 1 - \text{pbinom}(5, \text{size} = 9, \text{prob} = 1/2) \approx 0.2539$$
    
\item \textbf{Conclusion:} As the p-value is greater than 0.05, the data are consistent with the null hypothesis at the 5\% level of significance. There is no significant difference between the biochemical and the placebo.

\end{itemize}

3 negative, 6 positive:
```{r}
s = sign(rat$d)[sign(rat$d) != 0]
table(s)
```

Specify number of successes and the number of failures:
```{r}
binom.test(c(6, 3), p = 0.5, alternative = "greater")
```

Specify number of successes and the total number of trials:
```{r}
binom.test(sum(s==1), n = length(s), p = 0.5, alternative = "greater")
```
The p-value using the sign test is 0.254.

On the other hand, with paired $t$-test:
```{r}
t.test(rat$d, mu = 0, alternative = "greater")
```
The p-value using the $t$-test is 0.020! NOT TRUE

In this case the paired $t$-test and the sign test give conflicting results.

This is not uncommon when the sample size is small.
