---
title: "Live Lecture: ANOVA contrasts"
date: "2024-09-12"
author: "Ellen Ebdrup"
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---

```{r}
library(janitor)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

**Analysis of Variance**

# Plant Growth
The PlantGrowth data has results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions

We want to compare the means of the three different groups, the control, treatment 1 and treatment 2 groups.
```{r}
# the data is already "in" R
data("PlantGrowth") 

PlantGrowth |> ggplot() +
  aes(y = weight, x = group, 
      colour = group) + 
  geom_boxplot(coef = 10) + 
  geom_jitter(width = 0.1, size = 5) + 
  theme(legend.position = "none") +
  labs(y = "Weight (g)", x = "Group")
```

hasn't put in a total row, but you could add that yourself
F statistic: ratio between Mean Sq of 'group' and 'Residuals'
```{r}
plant_anova = aov(weight ~ group, data = PlantGrowth)
summary(plant_anova)
```
## Yield
```{r}
plant_summary = PlantGrowth |>
mutate(group = factor(group, levels = c("trt1","trt2", "ctrl"))) |>
group_by(group) |>
summarise(n = n(), mean_weight = mean(weight)) |>
mutate(contrast_coefficients = c(1,
-1, 0),
c_ybar = mean_weight * contrast_coefficients)
plant_summary
```
# two-sampl between treatment 1 and 2
```{r}
# estimated pairwise contrast:
sum(plant_summary$c_ybar) # sample contrast
```

## Residual standard error

```{r}
plant_anova = aov(weight ~ group, data = PlantGrowth)
summary(plant_anova)
```
$$\hat{\sigma}^2 = 0.3886$$
```{r}
library(broom)
# handy to convert ANOVA table to df
tidy(plant_anova) # makes a df st you can extract data with df$x
```

```{r}
resid_ms = tidy(plant_anova)$meansq[2]
resid_ms

resid_se = sqrt(resid_ms)
resid_se
```

### Calculating the test statistic

$$H_0 : \mu_1 - \mu_2 = 0$$
Finding the quantiles.
```{r}
qt(c(.025, .975), df=27) 
```


```{r}
(n_i = plant_summary |> pull(n)) # 10 10 10

(ybar_i = plant_summary |> pull(mean_weight)) # 4.661 5.526 5.032

(c_i = plant_summary |> pull(contrast_coefficients)) # 1 -1 0

(se = sqrt(resid_ms * sum((c_i^2) / n_i))) # 0.2787816

# observed test statistic
(t_stat = sum(c_i * ybar_i)/se) # -3.1
```

Reject or don't reject?

We would reject, since 
at a 5% significance, with a $t_{27}$-dist and anything outside [-2,2] you would rejct.
This means that the rejecting region is: 
$$RR = (-\infty , -2.05] \cup [2.05, \infty)$$  
### Calculating the p-value

```{r}
plant_anova$df.res

2*pt(abs(t_stat), df = plant_anova$df.res, lower.tail = FALSE)
```

### “Observed value” of confidence interval

=> Reject the $H_0$

```{r}
# the “multiplier” is determined via:
t_star = qt(0.975, df = 27)
t_star

# the interval is obtained using
sum(c_i * ybar_i) + c(-1,1) * t_star * se
```







