---
title: "Multiple testing"
date: "2024-09-09"
author: "Ellen Ebdrup"
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---

```{r}
library(janitor)
library(dplyr)
library(tidyverse)
library(ggplot2)
```


# microRNA and Alzheimer’s disease

```{r}
load("data/microRNA_full.RData")
```

```{r}
str(AD)
```

```{r}
# the 8 digit number is the subject id 
# the binary numbers indicate presence (1) or absence (0) of Alzheimer’s disease (AD)
head(AD)

# convert to df
disease_status = data.frame(AD) |> 
  tibble::rownames_to_column("subject")
str(disease_status)
```

```{r}
microRNA_Data[1:4,1:3]
```
Reshape the microRNA data and merge in the disease status information:
```{r}
mirna = microRNA_Data |>
  tibble::rownames_to_column("microRNA") |> 
  tidyr::pivot_longer(cols = -1, names_to = "subject", values_to = "value") |> 
  dplyr::left_join(disease_status)
head(mirna, n = 4)
```
## How many patients have Alzheimer’s?
```{r}
mirna |> select(subject, AD) |> 
  distinct() |> 
  janitor::tabyl(AD) |> 
  janitor::adorn_pct_formatting()

mirna |> 
  group_by(microRNA) |> 
  nest() |> 
  ungroup() |> 
  slice(1:15) |> # extract first 15 groups
  unnest(cols = everything()) |> 
  ggplot() + 
  aes(y = reorder(microRNA, value), 
      x = value, colour = factor(AD)) + 
  geom_boxplot(coef = 10) + 
  scale_color_brewer(palette = "Set1") + 
  theme(legend.position = "top") + 
  labs(colour = "Disease status",
       y = "MicroRNA")
```
## Welch 2-sample t-test
Let’s use Welch two-sample t-tests to compare the mean for people with Alzheimer’s to people without Alzheimer’s for all 309 microRNA.
```{r}
mirna_res = mirna |> 
  group_by(microRNA) |> 
  summarise(pval = t.test(value~AD)$p.value)
mirna_res |> ggplot() + aes(x = pval) + 
  geom_histogram(boundary = 0, binwidth = 0.05, 
                 fill = "skyblue",
                 colour = "black")

sum(mirna_res$pval < 0.05) # = 49
```
Of the 309 microRNA tested, 49 have p-values less than 0.05.

Are all of these “statistically significant” differences important?

- If there was no association between any microRNAs and Alzheimer’s disease our p-values follow a uniform distribution.
- We can generate a set of p-values knowing that there is no association and visualise this.

```{r}
set.seed(2)
mirna_res = mirna_res |> 
  mutate(null_pval = runif(n = n(), 
                           min = 0, 
                           max = 1))
mirna_res |> ggplot() + aes(x = null_pval) + 
  geom_histogram(boundary = 0, 
                 binwidth = 0.05,
                 fill = "midnightblue", 
                 colour = "black")

sum(mirna_res$null_pval < 0.05) # = 15
```

When we know that there are no truly important microRNAs, we still see 15 “significant” p-values in this simulated example.

# Error rates
## Controlling the family-wise error rate
$$FWEER=P(V\geq 1)$$

The probability of at least one false positive is $1-(1-\alpha)^m$

```{r}
m = 20
alpha = 0.05
1 - (1 - alpha)^m
```

### Bonferroni correction
  Pros: Easy to calculate, conservative
  Cons: May be very “conservative”
$$\alpha^* = \frac{\alpha}{m}$$
```{r}
m = 20
alpha = 0.05
1 - (1 - alpha/m)^m
```

**10 microRNA p-values: Bonferroni method**
```{r}
#install.packages("kableExtra")
alpha = 0.2
m = 10
M = nrow(mirna_res)
set.seed(123, sample.kind = "Rounding")
sample_rows = sample(1:M, size = m)
mirna10 = mirna_res |> 
  select(microRNA, pval) |> 
  slice(sample_rows) |> 
  mutate(p_bonferroni = pmin(pval*m, 1)) |> 
  arrange(pval)

mirna10 |> 
  knitr::kable(digits = 4) |> 
  kableExtra::kable_styling(
    font_size = 16)
```

## Controlling the false discovery rate
$$FDR=E\left[ \frac{FP}{R}\right]$$
where FP is Type I error

### Benjamini–Hochberg procedure

**10 microRNA p-values: BH method**
```{r}
alpha = 0.2
m = 10
p_vals = sort(mirna10$pval)
# BH procedure
# j=1: smallest p-value < 1*alpha/m?
p_vals[1] < 1*alpha/m
```

```{r}
# j=2: 
# second smallest p-value < 2*alpha/m?
p_vals[2] < 2*alpha/m
```

```{r}
# j=3: 
# third smallest p-value < 3*alpha/m?
p_vals[3] < 3*alpha/m
```


```{r}
# j=4: 
# fourth smallest p-value < 4*alpha/m?
p_vals[4] < 4*alpha/m
```

```{r}
# j=5: 
# fifth smallest p-value < 5*alpha/m?
p_vals[5] < 5*alpha/m
```

```{r}
# and so on ...
```

```{r}
# in general...
result = vector(length = length(p_vals))
p_vals = sort(p_vals) # we already did this but just emphasising it
for(j in seq(p_vals)) { # seq(p_vals) is the same as 1:length(pvals)
  result[j] = p_vals[j] < j*alpha/m
}
result
```

```{r}
largest_true = max(which(result == TRUE))
largest_true
```

```{r}
significant_pvals = p_vals[1:largest_true]
significant_pvals
```

**10 microRNA p-values: BH vs Bonferonni**
The lines are the significance thresholds for the three methods. If a point is below the line, the method would consider it “significant”.
```{r}
alpha = 0.2; m = 10
pvaldf = data.frame(p_vals) |> mutate(rank = rank(p_vals, ties.method = "random"))
pvaldf |> ggplot() + aes(y = p_vals, x = rank) + 
  geom_abline(aes(intercept = 0, slope=alpha*1/m, colour = 'Benjamini–Hochberg (FDR)'), linewidth = 1.5) +
  geom_hline(aes(yintercept = alpha, colour = 'No correction'), linewidth = 1.5) + 
  geom_hline(aes(yintercept = alpha/m, colour = 'Bonferroni (FWER)'), linewidth = 1.5) + 
  geom_point(size = 2) + 
  scale_x_continuous(breaks = 1:10) + 
  scale_colour_manual(name='Method',
                     breaks=c('No correction', 'Bonferroni (FWER)', 'Benjamini–Hochberg (FDR)'),
                     values=c('No correction'='blue',
                              'Bonferroni (FWER)'='red', 
                              'Benjamini–Hochberg (FDR)'='orange')) +
  theme_classic(base_size = 24) + labs(y = "p-value", x = "Ordering")
```


# Simulation experiments

## Case study I: no true positive
```{r}
set.seed(1234)
p_vals = rep(NA, 1000)
B = 10000
case1 = tibble(experiment = 1:B) |> 
  group_by(experiment) |> 
  reframe(x_sample = rnorm(20),
          y_sample = rnorm(20)) |> 
  group_by(experiment) |> 
  nest() |> 
  mutate(
    test = map(data, 
               ~t.test(.$x_sample, 
                       .$y_sample,
                       var.equal = TRUE) |> 
                 broom::tidy())) |> 
  unnest(test) |> 
  ungroup()
mean(case1$p.value < 0.05)
```

Get R to do the corrections for us using the p.adjust() function:
```{r}
case1 = case1 |> 
  mutate(
    p_bonf = p.adjust(p.value, 
                      "bonferroni"),
    p_bh = p.adjust(p.value, "BH")
  )
case1 |> select(experiment, p.value,
                 p_bonf, p_bh) |> 
  head()
```

Proportion of “significant” results
```{r}
case1 |> ungroup() |> 
  summarise(
    original = mean(p.value < 0.05),
    bonferroni = mean(p_bonf < 0.05),
    bh = mean(p_bh < 0.05)
  )
```


## Case study II: 50% true positives
```{r}
set.seed(1234)
B = 10000
case2 = tibble(experiment = 1:B) |> 
  group_by(experiment) |> 
  reframe(x_sample = rnorm(20), y_sample = rnorm(20)) |> 
  group_by(experiment) |> 
  rowwise() |> 
  mutate(truth = if_else(experiment<=B/2, "mu1 - mu2 = 0", "mu1 - mu2 = 2"),
         y_sample = if_else(truth == "mu1 - mu2 = 2", y_sample + 2, y_sample)) |> 
  ungroup() |> 
  nest(data = c(x_sample, y_sample)) |> 
  mutate(test = map(data, ~t.test(.$x_sample, .$y_sample, var.equal = TRUE) |> 
                      broom::tidy())) |> 
  unnest(test) |> ungroup() |> 
  mutate(
    prediction = if_else(p.value < 0.05, "reject H0", "don't reject H0"),
    p_bonf = p.adjust(p.value, method = "bonferroni"),
    p_bh = p.adjust(p.value, method = "BH"),
    pred_bonf = if_else(p_bonf < 0.05, "reject H0", "don't reject H0"),
    pred_bh = if_else(p_bh < 0.05, "reject H0", "don't reject H0"))
```


```{r}
# 1. row: no adjustment
# we cooretly rejected all the false H0
case2 |> tabyl(prediction, truth) |>
  gt::gt()

# 2. row: Bonferroni: controls FWER 
# never made any incorrect rejectings
case2 |> tabyl(pred_bonf, truth) |>
  gt::gt()

# 3. row: BH: controls FDR 
# we cooretly rejected all the false H0
# rejected some true H0, when we shouldnot have. 
case2 |> tabyl(pred_bh, truth) |>
  gt::gt()
```

```{r}
pval2 = case2 |> select(experiment, p.value, p_bonf, p_bh) |> 
  pivot_longer(cols = c(p.value, p_bonf, p_bh), 
               names_to = "method", 
               values_to = "p_value") |> 
  mutate(method = recode(method, 
                         "p.value" = "Original", 
                         "p_bh" = "BH", 
                         "p_bonf" = "Bonferroni"))
pval2 |> ggplot() + 
  aes(x = p_value, fill = method) + 
  geom_histogram(boundary = 0, binwidth = 0.05, colour = "black") + 
  facet_grid(~method) + 
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(breaks = c(0,1)) + 
  theme(legend.position = "none")
```


# MicroRNA revisited
```{r}
mirna_res = mirna_res |> 
  mutate(
    p_bonf = p.adjust(pval, method = "bonferroni"),
    p_bh = p.adjust(pval, method = "BH")
  )
mirna_res |> 
  summarise(original_n_sig = sum(pval < 0.05),
            bonf_n_sig = sum(p_bonf < 0.05),
            bh_n_sig = sum(p_bh < 0.05))
```

```{r}
library(gt)
mirna_res |> arrange(pval) |> 
  select(-null_pval, `Original p-value` = pval, 
         `Bonferroni p-value` = p_bonf,  `BH p-value` = p_bh) |> 
  head(n = 10) |> 
  gt() |> fmt_scientific(columns = 2:4) |> 
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")),
    locations = list(
      cells_body(columns = `Original p-value`, 
                 rows = `Original p-value` < 0.05),
      cells_body(columns = `Bonferroni p-value`, 
                 rows = `Bonferroni p-value` < 0.05),
      cells_body(columns = `BH p-value`, rows = `BH p-value` < 0.05)
    ))
```

Final comments
- Multiple testing is an entire subfield of statistics
- A basic Bonferroni/BH correction is often sufficient
- If there is strong dependence between tests there may be problems
