---
title: "PR21: Nearest neighbours"
output: html_document
date: "2024-10-21"
---

# Nearest neighbours

## Microchip test data
```{r}
library(tidyverse)
data = read_csv("https://raw.githubusercontent.com/DATA2002/data/master/Microchips.csv")
glimpse(data)
```


```{r}
ggplot(data) +
  aes(x=Test1, y=Test2, 
      color=factor(Label)) + 
  geom_point(size=5, alpha = 0.8)
```

## kNN

### kNN in R

Split our data into predictors (X), and response (y),
```{r}
library(class)
X = data |> 
  dplyr::select(Test1, Test2)
y = as.factor(data$Label)
```


Use knn to predict a grid of new data points:

```{r}
new.X = expand.grid(
  Test1 = seq(min(X$Test1 - 0.5),
              max(X$Test1 + 0.5), 
              by = 0.1),
  Test2 = seq(min(X$Test2 - 0.5), 
              max(X$Test2 + 0.5),
              by = 0.1)
  )
```

```{r}
p = ggplot(new.X) +
  geom_point(aes(x=Test1, y=Test2), 
             alpha=0.1)
p
```

Now we call the knn function, putting in our original observations and their observed the classes as well as the grid of “new points” that we’ll be predicting.
We also need to specify how many neighbours we want to learn from.
For now let $k=3$.

```{r}
pred_k3 = knn(train = X, test = new.X, 
              cl = y, k = 3, prob = TRUE)
pred_prob_k3 = attr(pred_k3, "prob")
```

We can also see how our decisions change based on the number of neighbours we consider.
Here, we consider k=9.

```{r}
pred_k9 = knn(train = X, test = new.X, 
              cl = y, k = 9, prob = TRUE)
pred_prob_k9 = attr(pred_k9, "prob")
```

## Performance assessment

Let’s compare k=3 and k=9
In sample confusion matrix

```{r}
library(caret)

# k=3
k3 = knn(train = X, test = X, cl = y, k = 3)
confusionMatrix(k3, y)$table
confusionMatrix(k3, y)$overall[1] |> round(2)

#k=9
k9 = knn(train = X, test = X, cl = y, k = 9)
confusionMatrix(k9, y)$table
confusionMatrix(k9, y)$overall[1] |> round(2)
```


```{r}
res = data.frame(
  k = 1:20,
  accuracy = NA
)
for(k in res$k){
  k_temp = knn(train = X, test = X, 
               cl = y, k = k)
  cmat = confusionMatrix(k_temp, y)
  res$accuracy[k] = cmat$overall[1]
}
```


```{r}
ggplot(res) + 
  aes(x = k, y = accuracy) +
  geom_point() + geom_line() 
```


```{r}

```



```{r}

```


```{r}

```


```{r}

```



```{r}

```




