---
title: "Week 1 Lab 00."
date: "2024-08-01"
author: "Ellen Ebdrup"
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---
# install tidyverse, palmerpenguins and janitor

Load libraries
```{r}
library("tidyverse")
library("palmerpenguins")
```

Get a glimpse of the penguins data set
```{r}
glimpse(penguins_raw)
```
Clean the column names using janitor.clean_names()

```{r}
old_names = colnames(penguins_raw)
penguins = penguins_raw %>%
    janitor::clean_names()
```

```{r}
new_names = colnames(penguins)
dplyr::bind_cols(original = old_names, clean = new_names)
```

Be aware that the aes() is the new cleaned column names 
geom_point() adds the points 

```{r}
library("ggplot2")
penguins %>% 
  ggplot() + 
  # add the aesthetics
  aes(x = body_mass_g, 
      y = flipper_length_mm,
      colour = species) +
  # add a geometry
  geom_point() + 
  # tidy up the labels
  labs(x = "Body mass (g)",
       y = "Flipper length (mm)",
       colour = "Species")
```



```{r}
penguins = penguins |>
    mutate(species = stringr::word(species, start = 1, end = 1))
library("ggplot2")
penguins %>% 
  ggplot() + 
  # add the aesthetics
  aes(x = body_mass_g, 
      y = flipper_length_mm,
      colour = species) +
  # add a geometry
  geom_point() + 
  # tidy up the labels
  labs(x = "Body mass (g)",
       y = "Flipper length (mm)",
       colour = "Species")

ggsave(filename = "myfirstggplot.png")
```


## 2.2.1 Exercises

1. Generate a scatter plot for another pair of (numeric) variables.
2. Colour by sex and use facet_wrap() to generate a plot for each species and island combination.
3. Try including a line of best fit by adding another geometry layer geom_smooth(method = "lm").

```{r}
penguins %>% 
  drop_na(sex) %>%
  ggplot() + 
  
  # add the aesthetics
  aes(x = culmen_length_mm, 
      y = flipper_length_mm,
      colour = sex) +
  
  # add a geometry
  geom_point(alpha = 0.5) + 
  facet_wrap(vars(species, island)) +
  geom_smooth(method = "lm", se = FALSE) +
  
  # tidy up the labels
  labs(x = "Culmen length (mm)",
       y = "Flipper length (mm)",
       colour = "Sex") +
  scale_color_brewer(palette = "Set1") + 
  theme(legend.position = "bottom")

ggsave(filename = "my_awesme_plot.png")
readr::write_csv(penguins, file = "penguins_cleaned.csv")
```

4. Use a different geometry, geom_histogram() to create a histogram for flipper length, coloured by species.

```{r}
penguins %>% 
  drop_na(sex) %>%
  ggplot() + 
  
  # add the aesthetics
  aes(x = flipper_length_mm,
      fill = sex) +
  geom_histogram() +
  labs(x = "Flipper length (mm)",
       colour = "Sex") +
  scale_color_brewer(palette = "Set1") + 
  theme(legend.position = "bottom")
```
```{r}
# install.packages("plotly")
library("plotly")
myplot = penguins |> 
  ggplot() +
  # add the aesthetics
  aes(x = body_mass_g,
      y = flipper_length_mm,
      colour = species) +
  # add a geometry
  geom_point() +
  # tidy up the labels
  labs(x = "Body mass (g)",
       y = "Flipper length (mm)",
       colour = "Species")
plotly::ggplotly(myplot)
```


