---
title: "Lab 02-Smoking rates"
date: "2024-08-08"
author: "Ellen Ebdrup"
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---

A study of patients with insulin-dependent diabetes was conducted to investigate the effects of cigarette smoking on renal and retinal complications. 

```{r}
y_i = c(44, 24, 13,  19)
p_i = c(0.5, 0.2, 0.1, 0.2)
n = sum(y_i)
e_i = n * p_i
```

significance = 0.01

Should the researcher revise his estimates? 
Use 0.01 as the level of significance.

```{r}
t0 = sum((y_i - e_i)^2 / e_i)
t0

p_value = pchisq(t0, df=3, lower.tail = FALSE)
p_value 
```

Simpler:
```{r}
chisq.test(y_i, p=p_i)
```
1. Hypothesis 

$H_0 : p_1 =0.5, p_2=0.2, p_3=0.1, p_4=0.2$ vs $H_1$ : At least one of the equalities doesn't hold.

2. Assumption
Observations are randomly selected & independent of each other and $e_i = np_i \geq 5 \forall i$. 

3. Test Statistic
$$T=\sum_{i=1}^4 \frac{(Y_i - e_i)^2}{e_i} \sim x_3^2 \text{ Under } H_0$$

4. Observed Test Statistic
$$t_0=\sum_{i=1}^4 \frac{(y_i - e_i)^2}{e_i} \sim x_3^2 \text{ Under } H_0 = 2.47$$

5. P-value
$$P(T \geq t_0) = P( x_3^2 \geq 2.47) = 0.481$$

6. Decision
Since the p-value is greater than $\alpha=0.05$ (2.47 > 0.05), we don’t reject $H_0$, meaning that the data is consistent with the proportions estimated by the researcher.
