---
title: "Lecture: Collecting data"
date: "2024-08-19"
author: "Ellen Ebdrup"
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---

## Initial results (survival over a 20 year period)

```{r}
library(tidyverse)
x = read_csv("data/appleton1996.csv")
x
```
is called a tibble: 4 x 3


```{r}
x_long = tidyr::uncount(x, weights = count)
dim(x_long)
```

```{r}
x_long %>% dplyr::group_by(status) %>%
  dplyr::summarise(
    rate = sum(survival == "Died")/n()
)
```
is called a tibble: 2 x 2

```{r}
x_long %>% # without group_by()
  dplyr::summarise(
    rate = sum(survival == "Died")/n()
)
```

```{r}
ggplot(x) +
  aes(x = status,
      y = count,
      fill = survival) +
  geom_bar(stat = "identity",
           position = "fill") +
  scale_y_continuous(
    labels = scales::percent_format()) +
  labs(x = "", y = "Proportion") +
  theme_bw(base_size = 10) +
  scale_fill_brewer(palette = "Set1")
```

## Mortality by age group

```{r}
y = read_csv("data/appleton1996_age.csv")
dplyr::glimpse(y, width = 40)
```
Some conversion for easier read
```{r}
ytab = y %>%
  tidyr::pivot_wider(
    id_cols = age_group,
    names_from = c(status, survival),
    values_from = count,
    names_sep = " ")
ytab
```

```{r}
mortality = y %>% 
  uncount(weights = count) %>% 
  group_by(status, age_group) %>% 
  summarise(rate = mean(survival=="Died"))

p = mortality %>% 
  ggplot() + 
  aes(x = age_group, y = rate, fill = status) + 
  geom_bar(stat = "identity", 
           position = "dodge") + 
  theme_minimal(base_size=12) + 
  scale_fill_brewer(palette = "Paired") + 
  scale_y_continuous(
    labels = scales::percent_format()) +
  labs(title = "Mortality rates by age group",
       y = "Mortality rate", 
       x = "Age group", 
       fill = "") + 
  theme(panel.grid.major.y = element_blank(), 
        legend.position = "bottom") + 
  coord_flip()
p
```
Notice: 
- Not many young people died.
- Most old people died.
- In the middle age groups, smokers tended to have higher mortality rates than nonsmokers.

```{r}
p2 = y %>% 
  uncount(weights = count) %>%
  ggplot() + 
  aes(x = age_group, fill = status) + 
  geom_bar() + 
  theme_minimal(base_size = 12) + 
  scale_fill_brewer(palette = "Paired") + 
  labs(y = "Count", 
       x = "Age group", 
       fill = "") + 
  theme(panel.grid.major.y = element_blank(), 
        legend.position = "bottom") + 
  guides(fill = guide_legend(reverse = TRUE)) + 
  coord_flip()
p2
```
As there are many more young women who smoked than older women, and as younger
women are expected to live longer than older women, adding all the groups together makes smoking appear to be beneficial.

This is a classic example of Simpson’s paradox: 
  a trend present within multiple groups can reverse when the groups are     combined.
