---
title: "Lecture: Testing for homogeneity"
output: html_document
date: "2024-08-12"
---

```{r}
library(tidyverse)
dat = read_csv("https://raw.githubusercontent.com/DATA2002/data/master/covidplasma.csv")
dplyr::glimpse(dat)
```

Notice that "subject" should be an integer and not a double type representing each patient.

```{r}
dat = dat %>% 
  # remove censored data
  filter(outcome != "Censored") %>%
  
  # convert from a character type level to a factor type level
  mutate(treatment = factor(treatment, levels = c("Plasma", "No plasma")),
           outcome = factor(outcome, levels = c("Died", "Discharged")))
```

```{r}
table(dat$treatment, dat$outcome)
```

Prints a nice table

```{r}
dat %>%
  janitor::tabyl(treatment, outcome) |>
  gt::gt()
```



```{r}
# by default R does some corrections in the chisq.test. But we want the real asweer, therefore add the second argument. 
chisq.test(tab, correct = FALSE)
```
## Proces

1.  Hypothesis
$H_0 : p_{11} = p_{21}$ and $p_{12} = p_{22}$ vs $H_1 : p_{11} \neq p_{21}$ and $p_{12} \neq p_{22}$

2.  Assumption Assumptions: observations randomly sampled from two independent populations and $e_{ij} = \frac{y_{i \cdot} y_{\cdot j}}{n} \geq 5$.

3.  Test Statistic 
$$T = \sum_{i=1}^2 \sum_{i=j}^2 \frac{(Y_{ij} - e_{ij})^2}{e_{ij}} \,\ \,\  T \sim x_1^2 \text{ Under } H_0$$

4.  Observed Test Statistic 
$$t_0=\sum_{i=1}^2 \sum_{i=j}^2 \frac{(y_{ij} - e_{ij})^2}{e_{ij}} $$

5.  P-value 
$$P(T \geq t_0) = P( x_1^2 \geq t_0)$$

6.  Decision
Since the p-value is 0.1629 > 0.05, we don't reject the $H_0$ hypothesis.


# in general tables
```{r}
y = c(62, 47, 29, 46, 9, 7)
n = sum(y)
c = 3
r = 2
tab = matrix(y, nrow = r, ncol = c)  
# default is to fill by column
colnames(tab) = c("Approve ", 
                  " Not approve ",
                  " No comment")
rownames(tab) = c("Labor", "Liberal")
tab
```

```{r}
chisq.test(tab, correct = FALSE)
```


## Proces

1.  Hypothesis
$H_0 : p_{1j} = p_{2j} = ... = p_{rj}$ for $j=1,2...,c$ vs $H_1$ : Not all equalities hold.

or 

$H_0$ : the proportions of approve, not approve and no comment are homogenous across Liberal and Labour voters vs. $H_1$ : the proportions of approve, not approve and no comment are not the same across Liberal and Labour voters.

2.  Assumption 
Assumptions: $e_{ij} = \frac{y_{i \cdot} y_{\cdot j}}{n} \geq 5$ and independent observations sampled from the  populations

3.  Test Statistic $$T = \sum_{i=1}^r \sum_{i=j}^c \frac{(Y_{ij} - e_{ij})^2}{e_{ij}} \,\ \,\  T \sim x_{(r-1)(c-1)}^2 \text{ approx }$$

4.  Observed Test Statistic $$t_0=\sum_{i=1}^r \sum_{i=j}^c \frac{(y_{ij} - e_{ij})^2}{e_{ij}} $$

5.  P-value $$P(T \geq t_0) = P( x_{(r-1)(c-1)}^2 \geq t_0)$$

6.  Decision
The p-value is less than 0.05, therefore at the 5% level of significance, we reject the null hypothesis and conclude that voter preferences about the new tax reform package are not homogenous across Liberal and Labour voters.
