---
title: "Lecture: Testing for independence"
output: html_document
date: "2024-08-13"
---

## Prerecorded lecture

# Workflow: test of independence between two variables in a 2 x 2 table


1. Hypothesis:
$H_0$ : $p_{ij} = p_{i \bullet} p_{\bullet j}$ for $i=1,2$ and $j=1,2$ vs $H_1$ : Not all equalities hold 
or  
$H_0$ :variable 1 is independent of variable 2 vs $H_1$ : the two variables are not independent.

2. Assumptions: 
Independent observations and $e_{ij} = \frac{y_{i \bullet} y_{\bullet j}}{n} \geq 5$.

3. Test statistics:
$$T = \sum_{i=1}^2 \sum_{i=j}^2 \frac{(Y_{ij} - e_{ij})^2}{e_{ij}} \,\ \,\ . \text{ under } H_0 , T \sim x_1^2 \text{ approx }$$

4. Observed test statistic:
$$t_0=\sum_{i=1}^2 \sum_{i=j}^2 \frac{(y_{ij} - y_{i \bullet} y_{\bullet j} / n)^2}{y_{i \bullet} y_{\bullet j} / n} $$
5. P-value: 
 $$P(T \geq t_0) = P( x_1^2 \geq t_0)$$

6. Decision: 
Reject $H_0$ if the p-value $< \alpha$.



# Titanic: Adults in third class

$H_0$ : gender is independent of survival for adults in third class vs 
$H_1$ : gender and survival are not independent for adults in third class.

```{r}
titanic_df = as.data.frame(Titanic)
head(titanic_df)
```

Who and how many survived?
- Notice that a lot more men died compered to women. 
```{r}
y_mat = xtabs(Freq ~ Sex + Survived,
              data = titanic_df)
y_mat
```

Looking only at the 3rd class.
```{r}
t3a = titanic_df |> 
  filter(Class == "3rd",
         Age == "Adult")
y_mat = xtabs(Freq ~ Sex + Survived,
              data = t3a)
y_mat
```

```{r}
t3a |> 
  ggplot() + 
  aes(x = Sex, 
      y = Freq, 
      fill = Survived) + 
  geom_col() + 
  scale_fill_brewer(palette = "Set1")
```

```{r}
y_mat
```

```{r}
chisq.test(y_mat, 
           correct = FALSE) # so R doesnot do any corrections for us.
```
Result is a very small p-value, which means that we we reject the $H_0$ hypothesis.
We reject the null hypothesis that sex is independent of survival for adults in third class as the p-value is very small (much smaller than 0.05). 
Hence, there is evidence to suggest that survival status of passengers on the Titanic is related to the sex of the passenger.


# Testing for independence in general tables



