---
title: "Data2002 Project"
date: "`r Sys.Date()`"
author: ""
format: 
  html: 
    embed-resources: true # Creates a single HTML file as output
    code-fold: true # Code folding; allows you to show/hide code chunks
    code-tools: true # Includes a menu to download the code file 
table-of-contents: true # (Optional) Creates a table of contents
number-sections: true # (Optional) Puts numbers next to heading/subheadings
---


```{r}
# Maybe you need to run this:

#install.packages("paletteer")
#install.packages("qtlcharts")
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(janitor)
library(tidyverse)

# Plots
library(ggplot2)
library(gridExtra)
library(paletteer)
library(ggfortify)
library(qtlcharts) # Heatmap

# Model, analysis..
library(car)        
library(lmtest)     
library(MASS) 
library(leaps)
library(caret)
```
 


# Project: EDA

## Cleaning
```{r}
# Read in the datasets
math_df <- read.table("Data/student-mat.csv", sep=";", header=TRUE)
port_df <- read.table("Data/student-por.csv", sep=";", header=TRUE)

# Merge the datasets
df <- merge(port_df, math_df, by=c("school", "sex", "age", "address", "famsize", "Pstatus", 
                                   "Medu", "Fedu", "Mjob", "Fjob", "reason", "nursery", "internet"))
# Clean column names
df <- df %>% clean_names()

# Replace 'x' with 'math' and 'y' with 'port' in the column names
df <- df %>% rename_with(~ gsub("_x", "_math", .), everything()) %>%
              rename_with(~ gsub("_y", "_port", .), everything())
colnames(df)
```

## Histograms
```{r}
# Math - G3 distribution
p1 <- ggplot(df, aes(x = g3_math)) +
  geom_histogram(binwidth = 1, fill = "#C24841FF", color = "black") + 
  scale_fill_paletteer_d("nationalparkcolors::Acadia") +
  labs(title = "Final Grade in Mathematics", x = "Grade (G3_mat)", y = "Frequency")

# Portuguese - G3 distribution
p2 <- ggplot(df, aes(x = g3_port)) +
  geom_histogram(binwidth = 1, fill = "#476F84FF", color = "black") +
  labs(title = "Final Grade in Portuguese", x = "Grade (G3_port)", y = "Frequency")

grid.arrange(p1, p2, ncol = 2)
```
## Boxplot
```{r}
# Combine studytime for Math and Portuguese in one dataframe for boxplot
studytime_df <- df %>%
  select(studytime_math, studytime_port, g3_math, g3_port) %>%
  pivot_longer(cols = c("studytime_math", "studytime_port"), names_to = "Subject", values_to = "studytime")

# Combined boxplot for Study Time vs G3
ggplot(studytime_df, aes(x = factor(studytime), 
                         y = ifelse(Subject == "studytime_math", g3_math, g3_port), fill = Subject)) +
  geom_boxplot() +
  labs(title = "Final Grade (G3) vs. Study Time for both subjects", x = "Study Time", y = "Final Grade (G3)") +
  scale_fill_manual(values = c("studytime_math" = "#C24841FF", "studytime_port" = "#476F84FF"),
                    labels = c("studytime_math" = "Mathematics", "studytime_port" = "Portuguese")) + 
  theme_gray()
```
## Scatter plot with fitted linear line
```{r}
# Create a new dataframe with combined data for Math and Portuguese
combined_df <- df %>%
  mutate(subject = "Mathematics", g2 = g2_math, g3 = g3_math) %>%
  bind_rows(df %>% mutate(subject = "Portuguese", g2 = g2_port, g3 = g3_port))

# Scatter plot for G3 as a function of G2 for both Math and Portuguese in the same plot
ggplot(combined_df, aes(x = g2, y = g3, color = subject)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, aes(color = subject)) +
  
  # Custom colors for Math and Portuguese
  scale_color_manual(values = c("Mathematics" = "#C24841FF", "Portuguese" = "#476F84FF"),
                     labels = c("Mathematics", "Portuguese")) +
  
  # Labels and theme
  labs(title = "Final Grade (G3) vs. Second Period Grade (G2)",
       x = "Second Period Grade (G2)",
       y = "Final Grade (G3)",
       color = "Subject") +
  theme_grey()
```


# Project: Presentation

Go with ´math´
```{r}
df <- read.table("Data/student-mat.csv", sep=";", header=TRUE)
df <- df |> clean_names()
```


## Data exploration

```{r}
ggplot(df, aes(x = g3)) +
  geom_histogram(binwidth = 1, fill = "#476F84FF", color = "black") + 
  scale_fill_paletteer_d("nationalparkcolors::Acadia") +
  labs(title = "Final Grade (G3) in Mathematics", x = "Point scale from [0, 20]", y = "Frequency")
```


# Heapmap, numeric variables
```{r}
# Select numerical columns for correlation (e.g., grades, absences, age)
numerical_cols <- df[, sapply(df, is.numeric)]

# Generate the correlation matrix
cor_matrix <- cor(numerical_cols, use = "complete.obs")

# Create an interactive heatmap using qtlcharts
iplotCorr(cor_matrix)
```


## Initial full model 
```{r}
null_model <- lm(g3 ~ 1, data = df)
full_model <- lm(g3 ~ ., data = df)

summary(full_model)

# Diagnostic plots
autoplot(full_model, which = 1:3)
```


## Model selection

From **Thursday session**, we found that the signficant variables are: 
  ´sex´, ´mjob´, ´medu´, ´failures´, ´goout´

### Backward selection

```{r}
step.back.aic <- step(full_model, 
            direction = "backward", 
            trace = FALSE) # set to TRUE, if you want to print all model selection outputs
summary(step.back.aic)
``` 

Backward selection results:

g3 $\sim$ school + age + activities + romantic + famrel +  walc + absences + g1 + g2


### Forward selection
```{r}
# Perform forward selection
step.fwd.aic <- step(null_model, 
                  scope = list(lower = null_model, upper = full_model), 
                  direction = "forward",
                  trace = FALSE)
summary(step.fwd.aic)
```

Forward selection results:

g3 $\sim$ g2 + famrel + absences + g1 + age + activities +  walc + romantic + school

### Comparing
```{r}
sjPlot::tab_model(step.fwd.aic, step.back.aic, show.ci = FALSE, show.aic = TRUE,
                  dv.labels = c("Forward model", "Backward model"))
```

## Exhaustive search

```{r}
# Perform exhaustive search
exhaustive_search <- regsubsets(g3 ~ ., data = df, nvmax = ncol(df) - 1)
exhaustive_summary <- summary(exhaustive_search)

# Find the model with the smallest BIC
optimal_exhaustive_model <- which.min(exhaustive_summary$bic)
exhaustive_summary$which[optimal_exhaustive_model, ]
```

Exhaustive search result: 

G3 $\sim$ (Intercept) + age + famrel + absences + g1 + g2

## ´train_control´ function

```{r}
# Set seed for reproducibility
set.seed(123)

# Define control for cross-validation
train_control <- trainControl(method = "cv", 
                              number = 10,   # k = 10
                              repeats = 5)  
```

## CV of bwd/fwd selection model


### Own model


### Backward CV model
```{r}
# Fit the backward selection model
bwd_full_model_cv <- train(
  g3 ~ school + age + activities + romantic + famrel + walc + absences + g1 + g2, 
  data = df, 
  method = "lm", 
  trControl = train_control)

# Print results for backward selection model
bwd_full_model_cv
```

### Forward CV model
```{r}
# Fit the backward selection model
fwd_full_model_cv <- train(
  g3 ~ g2 + famrel + absences + g1 + age + activities + walc + romantic + school, 
  data = df, 
  method = "lm", 
  trControl = train_control)

# Print results for backward selection model
fwd_full_model_cv
```


### CV of exhaustive search model

```{r}
# Fit the model using cross-validation
cv_model <- train(g3 ~ g1 + g2 + famrel,
                  data = df, method = "lm", 
                  trControl = train_control)

# View the results
cv_model
```

```{r}
exh_leaps = regsubsets(g3 ~., data = df, nvmax = 10)
summary(exh_leaps)$outmat

exh = lmSubsets(g3 ~ ., data = df, nbest = 10)
plot(exh)
```



## New model
```{r}
df2 <- read.table("Data/student-mat.csv", sep=";", header=TRUE)
df2 <- df2 |> clean_names()

df2 <- subset(df, select = -c(g1, g2))
```

```{r}
null_model <- lm(g3 ~ 1, data = df2)
full_model <- lm(g3 ~ ., data = df2)
```

```{r}
# Perform backward selection
step.back.aic <- step(full_model, 
            direction = "backward", 
            trace = FALSE) # set to TRUE, if you want to print all model selection outputs
#summary(step.back.aic)
``` 

```{r}
# Perform forward selection
step.fwd.aic <- step(null_model, 
                  scope = list(lower = null_model, upper = full_model), 
                  direction = "forward",
                  trace = FALSE)
#summary(step.fwd.aic)
```


```{r}
sjPlot::tab_model(step.fwd.aic, step.back.aic, show.ci = FALSE, show.aic = TRUE,
                  dv.labels = c("Forward model", "Backward model"))
```

## Exhaustive search

```{r}
# Perform exhaustive search
exhaustive_search <- regsubsets(g3 ~ ., data = df2, nvmax = ncol(d2)-1)
exhaustive_summary <- summary(exhaustive_search)

# Find the model with the smallest BIC
optimal_exhaustive_model <- which.min(exhaustive_summary$bic)
exhaustive_summary$which[optimal_exhaustive_model, ]
```


```{r}
exh_leaps = regsubsets(g3 ~., data = df2, nvmax = ncol(d2)-1)
summary(exh_leaps)$outmat
```

```{r}
exh = lmSubsets(g3 ~ ., data = df2, nbest = ncol(d2)-1)
plot(exh)
```
