---
title: "Live lecture: Assessing performance"
format: html
---

# In sample performannce

## Fitted ozone model
```{r}
library(tidyverse)
data(environmental, package = "lattice")
environmental = environmental |> 
  mutate(lozone = log(ozone))
lm2 = lm(lozone ~ temperature, data = environmental)
lm3 = lm(lozone ~ radiation + temperature + wind, environmental)
```

## In sample performance vs out of sample performance

In sample: $r^2$, comparing the simple linear regression to the full model

```{r}
summary(lm2)$r.squared # lozone ~ temperature

summary(lm3)$r.squared # lozone ~ radiation + temperature + wind
```

Out of sample: How well do we predict observations that we didn’t use to build the model?

# Out of sample performance

## Comparing simple linear regression with the full model

Out of sample performance
We could think about building a training set and using it to predict observations from a test set.
```{r}
n = nrow(environmental)
n
```

```{r}
# this box is from the slides
n_train = floor(0.8*n)
n_test = n - n_train
grp_labs = rep(c("Train","Test"), times = c(n_train, n_test)) 
environmental$grp = sample(grp_labs)
train_dat = environmental |> filter(grp == "Train")
lm_simple_train = lm(lozone ~ temperature, data = train_dat)
lm_full_train = lm(lozone ~ radiation + temperature + wind, data = train_dat)
test_dat = environmental |> filter(grp == "Test")
simple_pred = predict(lm_simple_train, newdata = test_dat)
full_pred = predict(lm_full_train, newdata = test_dat)
```

Split the data
```{r}
n_train = floor(0.8*n) # 88
n_test = n - n_train # 23
```


```{r}
set.seed(1) # for reuse ability
grp_labs = rep(c("Train","Test"), times = c(n_train, n_test)) 
table(grp_labs)
# sample group labs without replacement
# st we achieve random sampling 
environmental$grp =  sample(grp_labs, replace = FALSE) 
```

See the environmental data
```{r}
environmental
```


```{r}
train_dat = environmental |> filter(grp == "Train") # take the train group
test_dat = environmental |> filter(grp == "Test")
```

fit our model on our training data:
```{r}
lm_simple_train = lm(lozone ~ temperature, data = train_dat)
lm_full_train = lm(lozone ~ radiation + temperature + wind, data = train_dat)
```


### Root mean squared error (RMSE)

Residuals: difference from what we predicted vs. true value using the same data set

How can we compare the predictions from the two models? Compare them to the observed values using the root mean square error:
$$RMSE = \sqrt{\frac{\sum_{i=1}^n (y_i-\hat{y}_i)^2}{n}}$$
$n$: number of train data points
$\hat{y}_i$ : predicted

$(y_i-\hat{y}_i)^2$ : squared prediction error

predict on out test data
```{r}
simple_pred = predict(lm_simple_train, newdata = test_dat)
full_pred = predict(lm_full_train, newdata = test_dat)
```

```{r}
simple_mse = mean((test_dat$lozone - simple_pred)^2)
sqrt(simple_mse)
```

```{r}
full_mse = mean((test_dat$lozone - full_pred)^2)
sqrt(full_mse)
```


### Mean absolute error
An alternative measure of performance, less influenced by outliers
- a bit more robust, since it is not squared
$$MAE = \frac{\sum_{i=1}^m | y_i-\hat{y}_i |}{m}$$
$m$: number of test data points
```{r}
simple_mae = mean(abs(test_dat$lozone - simple_pred))
simple_mae
```


```{r}
full_mae = mean(abs(test_dat$lozone - full_pred))
full_mae # = 0.47 
```
In the slide code, the `full_mae` is a little lower since there wasn't set a seed. 


# Cross validation

k-fold cross-validation (CV) estimation
-Data randomly divided into k subsets of (nearly) equal size
- Estimate your model by leaving one subset out
- Use your estimated model to predict the observations left out
- Compute error rates on the left out set
- Repeat k times (for each of the subsets)
- Average the error rate over the k runs

Bias-variance tradeoff: smaller k can give larger bias but smaller variance

Computationally intensive.

often called *leave one out* cross validation

## 10-fold cross validation

**Step 1**: divide our data up into 10 folds there are 111 observations, so we have 9 folds of 11 observations and 1 fold of 12 observations.
```{r}
set.seed(2)
nrow(environmental) # = 111
environmental$grp = NULL # remove the grp variable we added previously
fold_id = c(1, rep(1:10, each = 11))
table(fold_id)
```


```{r}
environmental$fold_id = sample(fold_id, replace = FALSE)
head(environmental)
```



```{r}
k = 10
simple_mse = full_mse = vector(mode = "numeric", length = k)
simple_mae = full_mae = vector(mode = "numeric", length = k)
```


```{r}
res = data.frame(pred = NA, lozone = NA)
for(i in 1:k) { 
  
  test_set = environmental[fold_id == i,]
  training_set = environmental[fold_id != i,]
  
  simple_lm = lm(lozone ~ temperature, data = training_set)
  simple_pred = predict(simple_lm, test_set)
  
  # what I tried to do in lectures almost worked except I mispelt pred as red:
  res = rbind(res, cbind(pred = simple_pred , lozone = test_set$lozone))
  
  simple_mse[i] = mean((test_set$lozone - simple_pred)^2)
  simple_mae[i] = mean(abs(test_set$lozone - simple_pred))
  
  full_lm = lm(lozone ~ radiation + temperature + wind, data = training_set)
  full_pred = predict(full_lm, test_set)
  full_mse[i] = mean((test_set$lozone - full_pred)^2)
  full_mae[i] = mean(abs(test_set$lozone - full_pred))
  
}
dim(res)
# res # a data frame with the original y and the values that we predict from CV
# https://stats.stackexchange.com/questions/230913/how-caret-calculates-r-squared
cor(res, use = "pairwise.complete.obs")^2
```


```{r}
simple_mse |> sqrt() |> round(2)
```

**Step 2**: estimate the model leaving one fold out, make predictions on the test set and calculate the error rate
```{r}
# box from lecture slides
k = 10
simple_mse = full_mse = vector(mode = "numeric", length = k)
simple_mae = full_mae = vector(mode = "numeric", length = k)
for(i in 1:k) { # go through each of the 10 folds
  test_set = environmental[fold_id == i,] # only those test id's = i
  training_set = environmental[fold_id != i,] # the other data points/rows
  simple_lm = lm(lozone ~ temperature, data = training_set)
  simple_pred = predict(simple_lm, test_set)
  simple_mse[i] = mean((test_set$lozone - simple_pred)^2)
  simple_mae[i] = mean(abs(test_set$lozone - simple_pred))
  full_lm = lm(lozone ~ radiation + temperature + wind, data = training_set)
  full_pred = predict(full_lm, test_set)
  full_mse[i] = mean((test_set$lozone - full_pred)^2)
  full_mae[i] = mean(abs(test_set$lozone - full_pred))
}
```
we end up with 10 errors (since we have 10 folds)

**Step 3**: aggregate the errors over the 10 folds
```{r}
cv_res = tibble(simple_mse, full_mse, 
                simple_mae, full_mae)
cv_res
```
```{r}
# Find the averages:
cv_sum_res = cv_res |> 
  summarise(
    across(.cols = everything(), 
           mean)
  )
cv_sum_res |> t()
```
Common to report the root mean square errors (RMSE):
```{r}
sqrt(cv_sum_res[,1:2])
```

We could visualise the error rates for each of the 10 folds:
```{r}
cv_res |> gather(key = "metric", value = "error") |> 
  separate(col = metric, into = c("model","metric")) |> 
  ggplot(aes(x = model, y = error)) + facet_wrap(~metric, scales = "free_y") + 
  geom_boxplot() 
```


## `caret` (Classification And REgression Training)


simple model:  temperature
full model: radiation + temperature + wind

"tell me how to do it easily"

```{r}
# box from lecture slides
library(caret) # works for all types of different prediction models

cv_full = train(
  lozone ~ radiation + temperature + wind, environmental,
  method = "lm",
  trControl = trainControl(
    method = "cv", number = 10, # 10-fold cv
    verboseIter = FALSE
  )
)
cv_full
```
the Rsquared parameter is not a in-sample 

This does the same as above, this is just a little simpler
```{r}
library(caret)
tr_ctrl = trainControl(
    method = "cv", 
    number = 10,
    verboseIter = FALSE
  )

cv_full = train(
  lozone ~ radiation + temperature + wind, 
  data =  environmental,
  method = "lm",
  trControl = tr_ctrl
)
cv_full
```

```{r}
cv_simple = train(
  lozone ~ temperature, 
  environmental,
  method = "lm",
  trControl = tr_ctrl # the training method definded before
)
cv_simple
```


```{r}
# take the train objects and visualise thier performance
results <- caret::resamples(list(simple = cv_simple,  full = cv_full))

ggplot(results, metric = "MAE") +
  labs(y = "MAE") 

ggplot(results, metric = "RMSE") +
  labs(y = "RMSE")

ggplot(results, metric = "Rsquared") +
  labs(y = "Rsquared") 
```
Reuslt: 

the full model performs better


## Repeated CV


```{r}
library(caret)
tr_ctrl = trainControl(
    method = "repeatedcv", number = 10, repeats = 10,
    verboseIter = FALSE
  )
cv_full = train(
  lozone ~ radiation + temperature + wind, environmental,
  method = "lm",
  trControl = tr_ctrl
)
cv_full
```

```{r}
cv_simple = train(
  lozone ~ temperature, 
  environmental,
  method = "lm",
  trControl = tr_ctrl
)
cv_simple
```


```{r}
results <- caret::resamples(list(simple = cv_simple,  full = cv_full))
ggplot(results, metric = "MAE") +
  labs(y = "MAE")
```