---
title: "Lab 02-Tablet Devices"
date: "2024-08-08"
author: "Ellen Ebdrup"
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

4.1 Tablet devises

Observed chi squared test

We always assume H0 is true. 

```{r}
y_i = c(102, 32, 12, 4) # actual value
p_i = c(0.69, 0.21, 0.07, 0.03) # H0 distribution
n = sum(y_i) # 150
e_i = n * p_i # expected values
```

combine the last two groups since there appear frequency values below 5.
```{r}
y_i = c(102, 32, 12 + 4) # actual value
p_i = c(0.69, 0.21, 0.07 + 0.03) # H0 distribution
n = sum(y_i) # 150
e_i = n * p_i # expected values
```

Use formula from slides
```{r}
t0 = sum((y_i - e_i)^2 / e_i)
```
How well does our observed data fit the expected data?

Since t0 = 0.0963, we will now find the p-value

if p-val is < 0.05 => reject
if p-val is > 0.05 => don't reject

```{r}
p_value = pchisq(t0, df=2, lower.tail = FALSE)
p_value 
```

Easier code: Same as above but much simpler
```{r}
chisq.test(y_i, p=p_i)
```

1. Hypothesis 

$H_0 : p_1 =0.69, p_2=0.21, p_3=0.1$ vs $H_1$ : At least one of the equalities doesn't hold.

2. Assumption
Observations are randomly selected & independent of each other and $e_i = np_i \geq 5 \forall i$. 

3. Test Statistic
$$T=\sum_{i=1}^3 \frac{(Y_i - e_i)^2}{e_i} \sim x_2^2 \text{ Under } H_0$$

4. Observed Test Statistic
$$t_0=\sum_{i=1}^3 \frac{(y_i - e_i)^2}{e_i} \sim x_2^2 \text{ Under } H_0 = 0.096$$

5. P-value
$$P(T \geq t_0) = P( x_2^2 \geq 0.0963) = 0.953$$

6. Decision
Since the p-value is greater than $\alpha=0.05$ (0.953 > 0.05), we don’t reject $H_0$, meaning that the data is consistent with the distribution of tablet devices in 2012.

