---
title: "Week 2"

format: 

  html:

    code-tools: true

    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

install.packages("mosaic")

library(mosaic)
```

\# Odds ratios in the wild

\- COVIDSurg Collaborative and GlobalSurg Collaborative (2021). Effects of pre-operative isolation on postoperative pulmonary complications after elective surgery: an international prospective cohort study. \_Anaesthesia.\_ DOI: \[10.1111/anae.15560\](https://associationofanaesthetists-publications.onlinelibrary.wiley.com/doi/full/10.1111/anae.15560)

```{r}
table2 = matrix(c(67993, 1386,
                  6877, 91,
                  10483, 206,
                  8910, 252), 
                  ncol = 2, byrow = TRUE)

colnames(table2) = c("No complications", "Postop complication(s)")

rownames(table2) = c("No isolation",
                     "<= 3 days", 
                     "4-7 days",
                     ">= 8 days")

table2
```

\## ≤ 3 days isolation

Relative to baseline

```{r}
e1 = table2[1:2,2:1]

e1

mosaic::oddsRatio(e1, verbose = TRUE)
```

Understanding this output:

Since "1" is NOT within the confidence interval (0.5243 \< OR \< 0.8037), there is some relation between the results/something going on. =\> Correlation between isolation and complication

In the mosaic package: R\^+ should me in row 2, whereas in the lectures we would normally have it the other way around.

R\^+ = ≤ 3 days : presence of the risk factor

R\^- = No isolation

```{r}
e1

rowsums = rowSums(e1)

# looking within row 1 (no isolation)
p1 = e1[1, 1]/rowsums[1] 

# looking within row 1 (<= 3 days isolation)
p2 = e1[2, 1]/rowsums[2]

o1 = p1/(1 - p1)
o2 = p2/(1 - p2)

RR = p2/p1
OR = o2/o1

RR

OR
```

So if we're interpreting this,

```{r}
# OR confidence interval
se = sqrt(sum(1/e1))
log_or_ci = log(OR) + c(-1,1)*qnorm(0.95)*se
or_ci = exp(log_or_ci)
or_ci
```

Is this a statistically significant result? At what level of significance?

\## 4-7 days isolation

Relative to baseline

```{r}
e2 = table2[c(1,3),2:1]
e2

mosaic::oddsRatio(e2, verbose = TRUE)
mosaic::relrisk(e2)
```

=\> We do NOT reject the H0, since 1 is within the confidence interval.

\## ≥ 8 days isolation

Relative to baseline

```{r}
e3 = table2[c(1,4),2:1]
e3

mosaic::oddsRatio(e3, verbose = TRUE)
mosaic::relrisk(e3)
```

=\> We do NOT reject the H0, since 1 is within the confidence interval.
