---
title: "Prerecorded lecture: Critical values, rejection regions and confidence intervals"
output: html_document
date: "2024-08-20"
---

# Critical values and confidence intervals
## Finding quantiles in R

In R, we get quantiles using the qDISTRIBUTION() range of functions, fx: 
    qt(p, df = k)       for t dist
    qnorm(p)            for normal dist
    qchisq(p, df = k)   for chi-squared dist

```{r}
qt(0.05, df = 5)
```

```{r}
qnorm(0.05)
```

## Beer example

```{r}
x = c(374.8, 375.0, 375.3, 374.8, 374.4, 374.9)

qt(0.975, df = 5) # = 2.57
qt(0.995, df = 5) # = 4.03
```

```{r}
# The sample mean:
xbar = mean(x)
xbar

# The standard error:
se = sd(x)/sqrt(6)
se

# The discrepancy from the “given value” 375:
discrep=abs(xbar-375)
discrep 
```

### Using qt()

$$\alpha = 0.05 : P(t_{n-1} \leq c) = 1- \frac{\alpha}{2} = 0.975$$
In r code, c would be
```{r}
qt(0.975, df = n-1)
```


$$\alpha = 0.01 : P(t_{n-1} \leq c) = 1- \frac{\alpha}{2} = 0.995$$
In r code, c would be
```{r}
qt(0.995, df = n-1)
```

For a 95% confidence interval for $\mu$ we thus choose $c$  via
```{r}
c_95 = qt(0.975, df = 5)
c_95
# giving
xbar + c(-1,1) * c_95 * se
```
Note that this includes the “special value” 375 and so is consistent with our 0.05 false-alarm rate test earlier.

For a 99% confidence interval for $\mu$ we thus choose $c$ via
```{r}
c_99 = qt(0.995, df = 5)
c_99
# giving
xbar + c(-1,1) * c_99 * se
```
As we’d expect, this CI is wider, and also includes 375.

### Using t.test()

```{r}
# The default:
t.test(x, mu = 375) 

# Setting conf.level = 0.99:
t.test(x, mu = 375, conf.level = 0.99)
```


## One sided, beer example

The 95% “upper confidence limit” is
```{r}
c_05 = qt(.95, df = 5)
xbar + c_05 * se
```
which gives the one-sided confidence interval
```{r}
c(-Inf, xbar + c_05 * se)
```
For 99%,
```{r}
c_01 = qt(.99, df = 5)
c(-Inf, xbar + c_01 * se)
```
These both include 375!

### Using t.test()

```{r}
# The default:
t.test(x, mu = 375, alternative = "less") 

# Setting conf.level = 0.99:
t.test(x, mu = 375, alternative = "less", conf.level = 0.99)
```
Using the (one sided) p-value in our level of confidence gives us a confidence interval “right on the edge”.
```{r}
t.test(x, mu = 375, alternative = "less", conf.level = 1 - 0.1589)
```



## Two sided, beer example

Using the (two sided) p-value in our level of confidence gives us a confidence interval “right on the edge”.
```{r}
t.test(x, mu = 375, conf.level = 1 - 0.3177)
```

# Rejection regions

## Smoking and blood platelet aggregation

```{r}
before = c(25, 25, 27, 44, 30, 67, 53, 53, 52, 60, 28)
after =  c(27, 29, 37, 36, 46, 82, 57, 80, 61, 59, 43)
df = data.frame(before, after, difference = after - before)
```
Is blood platelet aggregation affected by smoking?

```{r}
n = length(df$difference)
dbar = mean(df$difference)
s_d = sd(df$difference)
t0 = dbar/(s_d/sqrt(n))
c(n, dbar, s_d, t0) |>
  round(2)
```


```{r}
alpha = 0.05
qt(1-alpha/2, df = n - 1)
```

### Rejection region for sample mean

Rescale data. 
```{r}
before = c(25, 25, 27, 44, 30, 67, 53, 53, 52, 60, 28)
after =  c(27, 29, 37, 36, 46, 82, 57, 80, 61, 59, 43)
df = data.frame(before, after, difference = after-before)
s_d = sd(df$difference)
s_d
```

```{r}
n = nrow(df)
mu0 = 0
crit_val = qt(0.975, df = n-1)
crit_val
```

```{r}
rrlower = mu0 - crit_val * s_d / sqrt(n)
rrupper = mu0 + crit_val * s_d / sqrt(n)
c(rrlower, rrupper) |> round(2)
```




