---
title: "Prerecorded lecture: Testing means"
output: html_document
date: "2024-08-20"
---

# One sample

# Beer contents

```{r}
x = c(374.8, 375.0, 375.3, 374.8, 374.4, 374.9)

mean(x)
sd(x)
```

Is the mean beer content less than the 375 ml claimed on the label?
Mean is the think line in horizontal in the box below, which is 374.87 < 375.

```{r}
y = c(374.8, 375.0, 375.3, 374.8, 374.4, 374.9)

library("ggplot2")
df = data.frame(y)
set.seed(124)

p1 = ggplot(df, aes(x = "", y = y)) +
  geom_boxplot(alpha = 0.5, coef = 10) + 
  geom_dotplot(binaxis = 'y', 
               stackdir = 'center') + 
  # insert the hypothesised line
  geom_hline(yintercept = 375, 
             colour = "blue",
             linetype = "dashed") + 
  labs(y = "Beer volume (ml)", x = "") +
  theme_bw(base_size = 12) + 
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
p1
```

## Hypothesis

$H_0$ : $\mu = 375$ vs. $H_1$ : $\mu < 375$.

## Test statistic
$$T=\frac{\bar{X}-\mu_0}{S/ \sqrt{n}} \text{ Under } H_0, T \sim t_{n-1}$$

using the R t test
```{r}
t.test(x, mu=mu_, alternative = "less")
```
or "by hand":
```{r}
mu_ <- 375
mean_ <- mean(x)
sd_ <- sd(x)
n <- length(x)

t0 <- (mean_ - mu_)/(sd_/sqrt(n))
t0

pval = pt(t0, n - 1)
pval
```

# Two samples

# Smokers and blood platelet aggregation

This is a two sample test, since non_smokers and smokers and two different indpependent groups.
```{r}
non_smokers = c(25, 25, 27, 44, 30, 67, 
                53, 53, 52, 60, 28)
smokers =  c(27, 29, 37, 36, 46, 82, 
             57, 80, 61, 59, 43)

dat = data.frame(
  platelets = c(non_smokers, smokers),
  status = c(rep("Non smokers", 
                 length(non_smokers)),
             rep("Smokers", 
                 length(smokers)))
)
library(dplyr)
sum = dat %>% 
  group_by(status) %>% 
  summarise(Mean = mean(platelets),
            SD = sd(platelets), 
            n = n())
sum
```
A prettier way to display data
```{r}
knitr::kable(sum, 
             format = "html", 
             digits = 1)
```
Is the aggregation of blood platelets affected by smoking?



## Visualising blood platelet aggregation
```{r}
library(ggplot2)
ggplot(dat) + aes(x = status, y = platelets) +
  geom_boxplot() + 
  geom_jitter(width = 0.15, size = 3, colour = "blue") + 
  labs(x = "", y = "Blood platelet\naggregation")
```

## t test

```{r}
t.test(smokers, non_smokers,
       alternative = "two.sided",
       var.equal = TRUE) # default is FALSE
```
or "by hand"
```{r}
x = non_smokers 
z = smokers
n = length(x)
mean(x) 
mean(z)
```

```{r}
sd_p_sqrd = ((n-1)*(sd(x)^2) + (n-1)*(sd(z)^2)) / (n+n-2)
t0 = (mean(x) - mean(z)) / (sqrt(sd_p_sqrd) * sqrt(1/n + 1/n))
t0 # take the absolute value.
```

## p-value

We get the p-value from the t.test() code.
$$2P(t_{2n-2} \geq |t_0|) = 2P(t_{20} \geq 1.14) = 0.27$$
## Decision
Large p-value so the data are consistent with $H_0$. 
There does not appear to be evidence that blood platelet aggregation levels are different in smokers.

## Welch statistic
```{r}
t.test(smokers, non_smokers, alternative = "two.sided")
```
# Paired samples t-test
# Smoking and aggregation (paired)

Blood samples from 11 individuals before and after they smoked a cigarette are used to measure aggregation of blood platelets.
```{r}
before = c(25, 25, 27, 44, 30, 67, 53, 
           53, 52, 60, 28)
after =  c(27, 29, 37, 36, 46, 82, 57, 
           80, 61, 59, 43)
df = data.frame(before, after,
  difference = after - before)
df
```
Is the aggregation affected by smoking?
$\mu_0$ / the data should be centered around 0, if there is no difference on blood platelets depending on individuals smoked or not. 

```{r}
df %>% 
  summarise(across(.cols = c(before, after),
                   .fns = list(Mean = mean, 
                               SD = sd, 
                               n = length))) %>% 
  pivot_longer(cols = everything(),
               names_sep = "_",
               names_to = c("time", ".value"))
```

```{r}
p = ggplot(df) +
  aes(x = "", y = difference) + 
  geom_boxplot() +
  geom_dotplot(binaxis = "y", stackdir = "center") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = 'Difference in blood platelet levels')+
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p
```

## Test statistic

These two test are actually the same. 
R does this in the background:
The pairred t test is = one sample t test on the differences of the observations


```{r}
t.test(df$after, df$before, paired = TRUE)

t.test(df$difference)
```
## Decision
Small p-value so we reject the null hypothesis. There is evidence that blood platelet aggregation levels change after smoking.



