---
title: "Live Lecture: Testing in small samples"
output: html_document
date: "2024-08-15"
---

# Lady tasting tea

```{r}
truth = c("milk", "tea", "tea", "milk", "tea", "tea", "milk", "milk")
predicted = c("milk", "tea", "tea", "milk", "tea", "tea", "milk", "milk")
tea_mat = table(truth, predicted)
tea_mat
```

If p-value test is not acceptable, since the test set is so small. 

```{r}
chisq.test(tea_mat, correct = FALSE)
```
As seen from the contingency table: $y_{11} = 4$

```{r}
fisher.test(tea_mat)
```

```{r}
fisher.test(tea_mat, alternative = "greater") # correct in this setting
```
### Traditional chi-squared test
```{r}
chisq.test(tea_mat, correct = FALSE)
```

### Fisher’s exact test
```{r}
fisher.test(tea_mat)
```

### Yates’ continuity correction
```{r}
chisq.test(tea_mat, correct = TRUE)
```
## Result from tests
You shouldn't just choose the test with the smallest p-value (e.i. Chi-squared)

Pick the approch that best fit the data: Fisher's exact test
- since: small data set


# Cancer of the larynx

```{r}
y_mat = matrix(c(21, 15, 2, 3), ncol = 2)
colnames(y_mat) = c("Controlled", "Not controlled")
rownames(y_mat) = c("Surgery", "Radiation therapy")
y_mat
```


## Fisher's exact test

Is surgery more effective than radiation therapy?
```{r}
fisher.test(y_mat, alternative = "greater")
```

# Permutation testing

```{r}
galton.dat <- matrix(c(5, 4, 1, 12, 42, 14, 2, 15, 10), 3, 3)
rownames(galton.dat) = c("Arches-B", "Loops-B", "Whorls-B")
colnames(galton.dat) = c("Arches-A", "Loops-A", "Whorls-A")
galton.dat
```


```{r}
chisq.test(galton.dat)

chisq.test(galton.dat)$expected |> round(1)
```

## Monte Carlo simulation


```{r}
row_totals = rowSums(galton.dat)
col_totals = colSums(galton.dat)
B = 10000
set.seed(123)
x_list = r2dtable(n = B,
                  r = row_totals,
                  c = col_totals)
x_list[[1]]
```

```{r}
chisq.test(x_list[[1]])
```

```{r}
rnd.chisq = numeric(B) # initialise an empty vector

for (i in 1:B){ # loop over B iterations
  # each time save the test statistic
  rnd.chisq[i] = chisq.test(x_list[[i]])$statistic
  }
# what proportion of times did we observe a test statistic
# as or more extreme than what we observed?
sum(rnd.chisq >= 11.1699)/B
```

After you run the above cell:
```{r}
# par(cex = 1.8)
hist(rnd.chisq)
abline(v = 11.1699, col = "purple", lwd = 2)
axis(1, 11.1699, col.axis = "purple")
```


```{r}
chisq.test(galton.dat, simulate.p.value = TRUE)

chisq.test(galton.dat, simulate.p.value = TRUE, B = 10000)
```
