---
title: "Adapted from Lab 1B Mammograms"
date: "2024-08-23"
author: "Ellen Ebdrup"
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---

## Description

Suppose that among 100,000 women with negative mammograms, 20 will have breast cancer diagnosed within 2 years; and among 100 women with positive mammograms, 10 will have breast cancer diagnosed within 2 years. 

## Clinicians would like to know if there is a relationship between a positive or negative mammogram and developing breast cancer?

```{r}
x = matrix(c(10, 20, 90, 99980), ncol = 2)
colnames(x) = c("Breast cancer: yes", "Breast cancer: no")
rownames(x) = c("Mammogram: positive", "Mammogram: negative")
x
```


### (a) Is this a retrospective or a prospective study?

We classify the women based on them having the risk factor $\Rightarrow$ Prospective studies

### (b) Is it appropriate to use a relative risk to quantify the relationship between the risk factor (mammogram result) and disease (breast cancer)?  
If so calculate the relative risk.

$$RR = \frac{a(c+d)}{c(a+b)}$$
Using the code below, we have to take 1/result, since mosaic thinks the row order is reverse such that $R^-$ is first. 
Either you reverse the roworder or take the 1/x. 
```{r}
1/mosaic::relrisk(x)
# same as
RR = (10*(20+99980))/(20*(10+90))
```

$RR > 1$ : breast cancer is more likely to occur in the positive mammogram group.

Given the risk factor is NOT present, the probability of having the disease is very small. The nominator is very large:  
$P(D^+ | R^+)$ is 500 times as large as $P(D^+ | R^-)$


### (c) Calculate the odds of having breast cancer for positive vs negative mammograms.

Same as for the RR, take the 1/result since we haven't reversed the row order for the risk factor. 

```{r}
1/mosaic::oddsRatio(x)
# same as
y = x[c(2,1), ] # change row order
mosaic::oddsRatio(y, verbose = TRUE)
# same as
OR = (10*99980)/(90*20)
```

$OR > 1$ : increased risk of disease  

Conclusion: There is a relation between the mammogram and breast cancer diagnosis.

The estimated odds of having breast cancer for patients with positive mammograms is 555 times the estimated odds for those with negative mammograms. 


### (d) Calculate the standard error for the log odds-ratio.

```{r}
log_OR = log(OR)
log_OR
se_log_OR = sqrt(1/10 + 1/20 + 1/90 + 1/99980) # marginal error (standard deviation)
```


### (e) Calculate a 95% confidence interval for the odds-ratio.

$$\log OR ± 1.96 \cdot SE(\log OR)$$
If $OR$ is inside of the interval: Don't reject $H_0$ and the results are significant.

```{r}
q_ = qnorm(0.975) # 1.96

# one method:
log_ci = c(log(or) - q_ * se, log(or) + q_ * se) # [lower, upper]
log_ci

# another method:
lower_log_ci <- log_OR - q_ * se_log_OR
upper_log_ci <- log_OR + q_ * se_log_OR

c(exp(lower_log_ci), exp(upper_log_ci))
```

$$CI = 6.32 ± 1.96 \cdot 0.40 = [252.9 , 1219.9]$$


### (f) Is there evidence that there might be a relationship between mammogram test results and breast cancer diagnosis?

Conclusion: $OR=555 \in CI \Rightarrow$ There is a relation between the mammogram and breast cancer diagnosis.
