---
title:  "Adapted from Lab 1C TV violence" 
date:   "2024-08-25" 
author: "Ellen Ebdrup" 
output: 
  html_document: 
    ### IMPORTANT ###
    # self_contained: true # Creates a single HTML file as output
    code_folding: show # Code folding; allows you to show/hide code chunks
    ### USEFUL ###
    code_download: true # Includes a menu to download the code file
    ### OPTIONAL ###
    df_print: paged # Sets how dataframes are automatically printed
    theme: readable # Controls the font, colours, etc.
    toc: true # (Useful) Creates a table of contents!
    toc_float: true # table of contents at the side
    number_sections: false # (Optional) Puts numbers next to heading/subheadings
---

## Description

A study of the amount of violence viewed on television as it relates to the age of the viewer yields the results shown in the accompanying table for 81 people.

```{r}
x = matrix(c(8, 18, 12, 15, 21, 7), ncol = 3)
colnames(x) = c("16-34", "35-54", "55+")
rownames(x) = c("Low violence", "High violence")
x
```

```{r}
n = sum(x) # => Check!

(xr = apply(x, 1, sum)) # sum(row)

(xc = apply(x, 2, sum)) # sum(col)

# calculate expected values
(ex = xr %*% t(xc) / n) # t(x) => transpose of x

# calculate test statistic
(t0 = sum((x - ex)^2 / ex)) 
```

### (a) Which test is most appropriate in this scenario?

This scenario involves categorical data with two variables: 
  age group and 
  the amount of violence viewed. 
  
A Chi-squared test of independence is the most appropriate test to determine whether there is an association between the age group and the amount of violence viewed.


### (b) Write down the appropriate null and alternative hypotheses.

$H_0$ : There is no association between the age group and the amount of violence viewed (they are independent).

$H_1$ : There is an association between the age group and the amount of violence viewed (they are not independent).

### (c) What are the assumptions required for this test. Are they satisfied here?

The assumptions for a Chi-squared test of independence are:  
1. The data should be in the form of counts or frequencies.  
2. The categories should be mutually exclusive.  
3. The expected frequency in each cell should be at least 5.  

To check the assumptions, we will calculate the expected frequencies in each cell and see if they are all greater than or equal to 5.

```{r}
x > 5 # => Check!
```


### (d) What is the approximate distribution of the test statistic under the null hypothesis?


The test statistic for the Chi-squared test follows a Chi-squared distribution with degrees of freedom given by:
$$df=(n_{row}−1) × (n_{col}−1) = 1 \times 2 = 2$$


```{r}
(res = chisq.test(x))

# another way to find the expected values
res$expected

# test statistic
t0 = res$statistic
df = res$parameter
```

### (e) Write down an expression for the p-value.

$$P(\chi_{2}^2 \geq 11.169) = 0.00376$$
```{r}
res$p.value
```


### (f) What is your decision for the test and why?

Let the significance level, $\alpha = 0.05$.

Since the p-value is much smaller than $\alpha = 0.05 > 0.0038$, we reject the null hypothesis and conclude that there is an association between age group and the amount of violence viewed.
