---
title: "Wilcoxon rank-sum test"
format: 
  html:
    code-tools: true
    embed-resources: true
---

Provided for the live lecture and as example code. 

```{r, message=FALSE}
library(tidyverse)
```

## Yield

vector of observed variables
```{r}
A = c(32, 29, 35, 28)     # n_A = 4
B = c(27, 31, 26, 25, 30) # n_B = 5
# Then N = 9 
dat = data.frame(
  yield = c(A, B),
  method = c(rep("A", length(A)),
             rep("B", length(B)))
)
```


Another dataset frmo Quiz \#4
```{r}
A = c(1.83, 0.50, 1.62, 2.48, 1.68, 1.88, 1.55, 2.06, 1.30)
B = c(0.878, 0.647, 0.598, 2.05, 1.06, 1.29, 1.06, 3.14, 1.29)
dat = data.frame(
  yield = c(A, B),
  method = c(rep("A", length(A)),
             rep("B", length(B)))
)
```


```{r}
library(ggplot2)
ggplot(dat, aes(x = method, y = yield)) + 
  geom_boxplot() + 
  geom_point(size = 4, colour = "blue")
```


```{r}
dat = dat |> mutate(rank = rank(yield))
dat
```

```{r}
w_A = dat |> 
  filter(method == "A") |> 
  pull(rank) |> 
  sum()
w_A
```

```{r}
sum_dat = dat |>
  group_by(method) |> 
  summarise(n = n(),
            w = sum(rank))
sum_dat
```

```{r}
n_A = sum_dat |> 
  filter(method == "A") |> 
  pull(n)
n_B = sum_dat |> 
  filter(method == "B") |>
  pull(n)
# using the sums of the A sample
w_A = sum_dat |> 
  filter(method == "A") |> 
  pull(w)
ew_A = n_A * (n_A + n_B + 1)/2 
minw_A = n_A * (n_A + 1)/2 

# check what the expected value is
ew_A # = 20
# and 
minw_A # = 10
```

### Find the p-value
Now you want to find the p-value

If lower.tail=TRUE, then you would get w ≤ the observed value 

Interested in: $P(W \geq 26)$
where 26 is the sum of ranks associated with A

With out -1:
   # test(26 - 10 , 5 , 4 , FALSE) ==> P(W>16)P(W=17)P(W=18)
We - 1 since:
  # test(26 - 10 -1, 5 , 4 , FALSE) ==> P(W>15)P(W=16)P(W-17)
  we want the specific value of W=16, thats why it is included
         
          
```{r}
c(minw_A, w_A, ew_A) # w_A > ew_A

# looking in the upper tail, so use lower.tail = FALSE
2 * pwilcox(w_A - minw_A - 1, n_A, n_B, lower.tail = FALSE)
```

### What if we use the sums of the ranks of the B sample?



```{r}
sum_dat
# using the sums of the B sample
w_B = sum_dat |> 
  filter(method == "B") |> 
  pull(w)
ew_B = n_B * (n_B + n_A + 1)/2 
minw_B = n_B * (n_B + 1)/2 
c(minw_B, w_B, ew_B)
# now looking in the lower tail
2 * pwilcox(w_B - minw_B, n_B, n_A)
```


```{r}
wilcox.test(A, B, paired = FALSE) 

# linear regression test
wilcox.test(yield ~ method, data = dat)
```


```{r}
t.test(A, B) # t.test(yield ~ method, data = dat)
```


## Latent heat of fusion

two independent samples, whereas before it was paried. 

```{r}
A = c(79.98, 80.04, 80.02, 80.04, 80.03, 80.03, 80.04, 
      79.97, 80.05, 80.03, 80.02, 80.00, 80.02)
B = c(80.02, 79.94, 79.98, 79.97, 79.97, 80.03, 79.95, 
      79.97)
heat = data.frame(
  energy = c(A,B),
  method = rep(c("A","B"), c(length(A), length(B))))
```

To perform the Wilson test, you have to rank the observation

some of the ranks have fx 4.5 (decimal) => means there are ties
```{r}
heat = heat |> 
  dplyr::mutate(r = rank(energy))
# or equivalently
# heat$rank = rank(heat$energy)
heat |> arrange(r) 
```

```{r}
heat |>
  dplyr::group_by(method) |> 
  dplyr::summarise(
    w = sum(r), # observed test statistic = 180
    xbar = mean(energy),
    sd = sd(energy),
    n = n()
  ) |> 
  knitr::kable(format = "markdown", 
               digits = 3)

# heat
```

Slide 20:
Find the p-value $= P(Z > 2.7)$
```{r}
1-pnorm(2.7 , 0 , 1) # 0.00347
```



```{r}
ggplot(heat, aes(x = method, y = energy)) + 
  geom_boxplot() + 
  geom_dotplot(stackdir = "center",
               binaxis = "y") +
  theme_linedraw(base_size = 12) + 
  labs(y = "Heat of fusion (cal/g)",
       x = "Method")
```


```{r}
heat_sum = heat |>
  dplyr::group_by(method) |> 
  dplyr::summarise(
    w = sum(r),
    xbar = mean(energy),
    s = sd(energy),
    n = n()
  )
heat_sum
```

```{r}
na = heat_sum$n[heat_sum$method == "A"]
nb = heat_sum$n[heat_sum$method == "B"]
N = na + nb # total number of observations
c(na, nb, N)
```


```{r}
w = heat_sum$w[heat_sum$method == "A"]
EW = na * (N + 1)/2
c(w, EW)
```

```{r}
sumsqrank = sum(heat$r^2)
g = N * (N + 1)^2/4
varW = na * nb * (sumsqrank - g)/(N * (N - 1))
t0 = (w - EW)/sqrt(varW)
t0
```

```{r}
1 - pnorm(t0)
```

### Test using wilcox.test

quite rubox to outliers

```{r}
wilcox.test(A, B, alternative = 'greater', correct = FALSE)
```

```{r}
t.test(A, B, alternative = 'greater')
```


### Robustness


```{r}
# change the first value for the B method
heat1 = heat
heat1$energy[14] = 80.20 # instead of 80.02
# recalculate ranks
heat1 = heat1 |> dplyr::mutate( 
  r = rank(energy)
)
heat1
```

```{r}
heat1 |>
  dplyr::group_by(method) |> 
  dplyr::summarise(
    w = sum(r),
    Mean = mean(energy),
    SD = sd(energy),
    n = n()
  ) |> 
  knitr::kable(digits = 3)
```


```{r}
ggplot(heat1) + 
  aes(x = method, y = energy) + 
  geom_boxplot() + 
  geom_dotplot(stackdir = "center",
               binaxis = "y") +
  labs(y = "Heat of fusion (cal/g)",
       x = "Method")
```


```{r}
wilcox.test(energy ~ method, data = heat1, 
            alternative = 'greater',
            correct = FALSE)
wilcox.test(energy ~ method, data = heat, 
            alternative = 'greater',
            correct = FALSE)
```

```{r}
t.test(energy ~ method, data = heat1, alternative = 'greater')
t.test(energy ~ method, data = heat, alternative = 'greater')
```

## Notched boxplot
slide 26.

They do not overlap
Remember our $H_0$ : There is no overlap in data
```{r}
ggplot(heat, aes(x = method, y = energy)) + 
  geom_boxplot(notch = TRUE) + 
  geom_dotplot(stackdir = "center",
               binaxis = "y") +
  theme_linedraw(base_size = 12) + 
  labs(y = "Heat of fusion (cal/g)",
       x = "Method")
```
